Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-14 scagnostics.Rmd
notation table added

bibliography: bibliography.bib
output: rticles::springer_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{amssymb}
  - \usepackage{xcolor}
---

@@

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, descriptively named: "outlying", "skewed", "sparse", "clumpy", "striated", "convex", "skinny" and "stringy". The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behavior of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides 12 current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr] provides optimisation routines, and visualization.

@@

A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data set, computed by taking a $d$-dimensional projection of an $n\times p$ data matrix. Typically the definition is such that larger values of $f$ indicate a more interesting distribution of observations, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ variables will find the most interesting projections. This section describes the 12 indexes that are to be used to explore bivariate association. Some data pre-processing, including standardization, is advisable, prior to optimizing the PPI.

@@

Making a plot always involves some choice of scaling. When a scatterplot is made, effectively, albeit under the hood, the data is scaled into a range of $[a, b]$ (often $a=0, b=1$) on both axes to print it on a page or display in a graphics device window. The range deliminates page space within which to draw. The upshot is that the original data scale is standardized to the range and aspect ratio on the display space. It maybe that the original range of X is $[1, 1000]$ and Y is $[1, 1.6]$ but the display linearly warps this to $[0, 1]$ and $[0, 1]$, giving both variables equal visual weight.

@@

\textcolor{red}{\cite{JS87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall interval is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardizing the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardization procedure, it would most likely be unwelcome effects?}
-->

## Notation

## New index functions
\label{sec:indexDef}
Focusing now on $d=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developed for the selection of variable pairs that show interesting features in a scatter plot, but we generalist the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimizing over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:^[We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.]

>>>>

bibliography: bibliography.bib
output: rticles::springer_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{amssymb}
  - \usepackage{xcolor}
  - \usepackage{bm}
---

@@

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behavior of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr] provides optimisation routines, and visualization.

@@

A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data set, computed by taking a $d$-dimensional projection of an $n\times p$ data matrix. Typically the definition is such that larger values of $f$ indicate a more interesting distribution of observations, and therefore maximizing $f$ over all possible $d$ dimensional projections of a data set with $p>d$ variables will find the most interesting projections. This section describes the seven indexes that are to be used to explore bivariate association. Some data pre-processing, including standardization, is advisable, prior to optimizing the PPI.

@@

Making a plot always involves some choice of scaling. When a scatterplot is made, effectively, albeit under the hood, the data is scaled into a range of $[a, b]$ (often $a=0, b=1$) on both axes to print it on a page or display in a graphics device window. The range deliminates page space within which to draw. The upshot is that the original data scale is standardized to the range and aspect ratio on the display space. It maybe that the original range of one variable is $[1, 1000]$ and the other is $[1, 1.6]$ but the display linearly warps this to $[0, 1]$ and $[0, 1]$, say, giving both variables equal visual weight.

@@

\textcolor{red}{\cite{JS87} discusses centering and sphering of the data points. Centering is needed if index is not translational invariant (should not be the case for any index considered here). Arguments for sphering are that scale effects should be extracted, we don't want to reproduce information already obtained by PCA. To my mind rescaling to an overall interval is more appropriate here. Rotations between parameters should not affect the outcome, but will make interpretation more difficult. Standardizing the variance will remove information about the posterior variance. Overall we are interested in strong association only, if these structures were affected by the standardization procedure, it would most likely be unwelcome effects?}

## Notation
-->

\begin{table}[htp]
\caption{Summary of notation}
\centering
\begin{tabular}{|l|p{8.2cm}|} \hline
$p$ & be the number of variables in the data. For physics models the obervable space is higher dimension, say $q$, and the data examined is the fitted model space, typically less than 10. \\
$d (=2) $ & projection dimension. For studying physics models, typically $d=2$ and this is the focus for the index definition. \\
$n$ & Number of observations, which for the physics models is the number of fitted models being examined and compared. \\
&\\
$\bm{X}_{n\times p} = (\bm{X}_1, ..., \bm{X}_p)$ & $n\times p$-dimensional data matrix, where variables $\bm{X}_j$ may be scaled or standardised, and $\bm{X}$ may be sphered\\
$\bm{Y}_{n\times 2} = (\bm{Y}_1, ..., \bm{Y}_2)$ & projected data matrix. where $Y_j = (\alpha_1 X_1, ..., \alpha_p X_p$ \\
$\bm{F}_{p\times 2} = (\bm{\alpha}_1, ..., \bm{\alpha}_2)$ & orthonomal projection matrix\\ 
&\\
$H$ & Convex hull \\
$A$ & Alpha hull \\\hline
\end{tabular}
\label{notation}
\end{table}

## New projection pursuit index functions
\label{sec:indexDef}

Table \ref{notation} summarises the notation used for this section. This an overview of the functions that are converted into projection pursuit indexes. Full details of the functions can be found inthe original sources.

<!--
Focusing now on $d=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developed for the selection of variable pairs that show interesting features in a scatter plot, but we generalist the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimizing over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:^[We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.]
-->
